---
title: "RMS Titanic"
author: "Sascha Jalinous"
date: 2026-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>%
  glimpse()
```

**Observations**:

- Class, sex, age, survived, n

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- The total count according to wikipedia is 2208
- Are there any differences?
  - Yes the database is missing 7 people.
- If yes, what might account for those differences?
  - While attempting to figure out what could account for the differences, I accidentally ended up on a different Wikipedia page ([Sinking of the Titanic - Wikipedia](https://en.wikipedia.org/wiki/Sinking_of_the_Titanic#Casualties_and_survivors)). In the third line of the original article, there is a link attached to the statistic that 2,208 people were aboard, which I clicked. There is then a large table at the bottom that says total people was only 2,201. These conflicting numbers both come from Wikipedia so it could be that the database makers only took from this table they could have missed the 7 people mentioned in the RMS Titanic page.
  - The 2,208 number also does not say where it came from whereas the Sinking of the Titanic table says the 2,201 number comes from the British Board of Trade so it's possible discrepancies within record keeping over the years have introduced conflicting accounts for how many people were aboard the ship.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
# summary(df_titanic)
# df_titanic %>%
#   select(n, everything())
# test <- 
#   df_titanic %>%
#     filter(Survived == "Yes")
# 
# sex_vector = c(test %>% 
#     filter(Sex == "Female") %>%
#       select(n), 
#     test %>% 
#     filter(Sex == "Male") %>%
#       select(n))

## TASK: Visualize counts against `Class` and `Sex`
ggplot(
  data=df_titanic %>%
    filter(Survived == "Yes")
) +
  geom_bar(
    mapping = aes(
      x = Sex, # Sex,
      fill = Class, # Class,
      weight = n
    )
  )
```

**Observations**:

- Most of the male survivors were part of the crew, while the lowest amount of male survivors were in 2nd class.
- Most of the female survivors were in 1st class, while the lowest amount of female survivors was in the crew. This number is interesting because one thing that comes to mind is the question of just how many female crew members were present? I would expect there would be a majority of male crew members so it's possible that this lowest survivor count is skewed just by a limited amount of female crew in the first place, but I can not accurately say anything definitive based on this graph alone.
- In the female section, the amount of survivors in the 2nd and 3rd class are similar, whereas in the male section, there is more disparity between the survivors in those same categories. I can't make any inferences as to why because the numbers really don't tell me much else except just basic differences.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
ggplot(
  data=df_prop %>%
    filter(Survived == "Yes")
) +
  geom_bar(
    mapping = aes(
      x = Sex, # Sex,
      fill = Class, # Class,
      weight = Prop
    )
  )
```

**Observations**:

- There were a lot more women survivors than male survivors.
- Majority of survivors were in 1st and 2nd class with 3rd class having fewer survivors and the crew having the least survivors.
- The male crew had the least survivors overall.
- Is there anything *fishy* going on in your plot?
  - In the code for calculating the proportions, we group the data by Class, Sex, and Age, but then proceed to do nothing with Age besides add it to the data, which causes a weird change to the proportions that is not accounted for in the graph. The graph only plots Class and Sex and nowhere does it mention Age despite it being factored into the grouping.
  - The count for male crew survivors is so much lower compared to the raw counts because there are no children in the crew so no extra individuals are being added to the proportion.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
ggplot(
  data=df_prop %>%
    filter(Survived == "Yes")
) +
  geom_bar(
    mapping = aes(
      x = Sex, # Sex,
      fill = Class, # Class,
      weight = Prop
    )
  ) +
    facet_grid(rows = vars(Age))
```

**Observations**:

- The amount of male and female child survivors is roughly equal which makes sense because their age was prioritized, not their sex, which is different to the adults.
- The majority of survivors were female adults because their sex was prioritized over the male passengers who had much lower counts of survival in comparison.
- The smallest group of survivors overall were the ones in 3rd class which makes sense because they were the least prioritized.
- The fewest survivors in the adult male categories were those in 2nd class which is interesting because you would expect it to be 3rd class. I wonder what happened that is not visible in the data set that explains this.
- A relatively large proportion of the female crew survived which is probably because there were fewer female crew members to begin with.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - q4's graph had strange counts due to Age being factored in but not being graphed.
  - This new plot separates the survivors based on age which allows for a much clear visual and understanding about what the proportions actually mean.
  - The difference in female and male survivor proportions was not accurately being presented because the male children survivors were being included so it skewed the proportions and made it seem like the difference in male and female survivors was less than actuality.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
