---
title: "Gapminder"
author: "Sascha"
date: 2026-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder %>%
  glimpse()

gapminder %>%
  summary()
```

**Observations**:

- Write all variable names here:
  - Country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
  gapminder %>%
  pull(year) %>%
  max()
year_min <- 
  gapminder %>%
  pull(year) %>%
  min()

year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(
    aes(
      x = continent,
      y = gdpPercap
    )
  ) + 
  geom_boxplot() +
  scale_y_log10()
```

**Observations**:

- Oceania has the highest gdp while Africa has the lowest gdp.
- There are two outliers in the Americas and one in Asia.
- Oceania has extremely small quantiles.
- The outlier in Asia is a huge outlier compared to the gdp of everything else.

**Difficulties & Approaches**:

- The box plots were way too small originally which was fixed by making the y axis a log scale.
- It's difficult to know what countries the outliers are because they're just dots. I can see the next question says identify the outliers so I will look at the database and see which ones have the highest gdp.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>%
  filter(continent == "Asia") %>%
  arrange(desc(gdpPercap))

gapminder %>%
  filter(year == year_min) %>%
  filter(continent == "Americas") %>%
  arrange(desc(gdpPercap))
```

**Observations**:

- Identify the outlier countries from q2
  - Kuwait, United States, Canada

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  # drop_na() %>%
  #filter(year == max(year)) %>%
  filter(year == min(year) | year == max(year)) %>%
  # filter(!(year_max == FALSE & year_min == FALSE)) %>%
  # droplevels() %>%
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United States", "Kuwait", "Canada", "Haiti")),
    mapping = aes(color = country),
    size = 2
  ) +
  # facet_wrap(
  #   facets = vars(year == max(year), year == min(year))) +
  #   ~ year == year_max) +
  # facet_grid(
  #   rows = vars(year == max(year)), 
  #   cols = vars(year == min(year))
  # ) +
  facet_wrap(~ year) +
  scale_y_log10()
```

**Observations**:

- Oceania started out with a very low range gdp that was the highest, and maintained a very high gdp in 2007 with a bit wider of a range.
- As time has gone on, gdp for every continent has increased.
- A very low outlier was introduced in 2007 in the Americas (Haiti) that was not there when graphing the 1952 graph originally.
- In 1952 Asia had the widest range of gdp, which continued in 2007 and the range got even wider.
- Africa has the lowest median in both graphs, despite increasing in 2007.
- The continent with the median that rose the highest between the two years is Europe.
- Kuwait is an outlier in both years, but in 1952 it was much more drastically ahead of everyone else but by 2007 its had gdp fallen dramatically.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder %>%
  filter(year == year_min) %>%
  ggplot(
    aes(
      x = continent,
      y = lifeExp
    )
  ) + 
  geom_boxplot() +
  scale_y_log10() +
  stat_summary(fun = median,
             geom = "line",
             size = 1,
             aes(group = 1),
             col = "red")
```

- There is a lot of zig-zagging between the medians of the continents that shows the amount of disparity in the medians.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(year == year_max) %>%
  ggplot(
    aes(
      x = continent,
      y = lifeExp
    )
  ) + 
  geom_boxplot() +
  scale_y_log10() + 
  # geom_smooth(method = "lm", color="blue")
  stat_summary(fun = median,
               geom = "line",
               size = 1,
               aes(group = 1),
               col = "blue")
```

- There is less zig-zagging between the continents which shows that over time life expectancy for the Americas, Asia, Europe, and Oceania became very similar.
- Africa's life expectancy increased but there was still a large disparity between it and all of the other continents.

```{r q5-task3}
gapminder %>%
  group_by(year, continent) %>%
  summarize(pop = sum(pop)) %>%
  ggplot(aes(year, pop, fill = continent)) +
  geom_area()
```

- Asia has consistently had the biggest population.

- Europe's population has been slowly increasing over the span of this data set.

- Oceania went from an inconceivable speck to a slightly observable lines which shows the drastic difference in population size. I googled it and today Oceania has a population size of around 46 million according to Wikipedia so it is super small compared to every other continent.

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(
    ratio_gdp = max(gdpPercap) / min(gdpPercap)
  ) %>%
  ggplot(aes(year, ratio_gdp)) +
  geom_line() +
  geom_point()
```

This is a graph of the ratio of the highest/lowest gdp of all the countries over the years.

- There is a large spike in 1972 after a decline and then a sharp fall off afterward.

```{r}
gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot(aes(year, gdpPercap)) +
  geom_line() +
  geom_point()

gpd_guh <- 
  gapminder %>% filter( year == 1972) %>% 
  arrange(desc(gdpPercap)) 
gpd_guh
```

I looked at the data set to see what country had the highest gdp ratio and it didn't surprise me when it was Kuwait since Kuwait was the massive gdp outlier in 1952. Doing some research, this was most likely caused by a stock market boom due to oil revenue which eventually crashed the economy in 1977.

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(
    ratio_lifeexp = max(lifeExp) / min(lifeExp)
  ) %>%
  ggplot(aes(year, ratio_lifeexp)) +
  geom_line() +
  geom_point()
```

This is a graph showing the ratio of the highest/lowest life expectancy of all the countries over the years.

- There is a huge spike in 1992 which means that there was a huge disparity in ratio.

- This is a smaller but still noticeable spike in 1977 which means there was a large disparity in ratio again.

```{r}
gapminder %>%
  filter(country == "Rwanda") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  geom_point()

rwanda <-
  gapminder %>%
  filter(
    year == 1992) %>%
  arrange(desc(lifeExp))
rwanda
```

The large disparity in 1992 is due to the fact that Japan had the highest life expectancy with 79 years while Rwanda had a life expectancy of 23. Upon doing some research, this is most likely due to the Rwandan Genocide which started in 1992, ramped up in 1994 and decreased afterward which would explain the extremely low life expectancy during that time period.

```{r}
gapminder %>%
  filter(country == "Cambodia") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  geom_point()

cambodia <-
  gapminder %>%
  filter(
    year == 1977) %>%
  arrange(desc(lifeExp))
cambodia
```

The smaller disparity in 1977 is due to the life expectancy of Iceland being 76 years while the life expectancy of Cambodia was 31 years. Doing some research, this is most likely due to the Cambodian Genocide which started in 1975 and ended in 1979 which would account for the extreme drop in life expectancy.
